---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
  
## Loading data
  
The activity.csv file, which results from unzipping activity.zip, should be located in the working directory.
  
```{r}
  data <- read.csv("./activity.csv")
```

## What is the mean total number of steps taken per day?

Missing value will be ignored when calculating the total number of steps taken per day.
Note that the library *dplyr* will be used in the rest of this document.

```{r stepsPerDay, message = FALSE}
library(dplyr)
stepsPerDay <- data[!is.na(data$steps),c("steps","date")] %>% group_by(date) %>% summarise_each(funs(sum))
stepsPerDay
```

The histogram of total number of steps taken per day :

```{r histogramStepsPerDay}
hist(stepsPerDay$steps, 
     breaks = 10,
     main = "Histogram of total number of steps taken each day",
     xlab = "Total number of steps taken on a day",
     ylab = "Number of cases")
```

The mean and median of total number of steps taken per day is calculated as follows:

```{r}
meanStepsPerDay <- mean(stepsPerDay$steps)
medianStepsPerDay <- median(stepsPerDay$steps)
```

The mean is `r meanStepsPerDay`  
The median is `r medianStepsPerDay`

##Average daily activity pattern

The average daily activity pattern can be visualized on a graph.   In order to identify the pattern, the approach taken here is to average the number of steps per interval over days.

```{r activityPattern}
stepsPerInterval <- data[!is.na(data$steps),c("steps","interval")] %>% group_by(interval) %>% summarise_each(funs(mean))
plot(stepsPerInterval$interval,stepsPerInterval$steps,
     type = 'l',
     main = "Steps per interval, averaged over days",
     xlab = "Interval",
     ylab = "Steps")
```

Let's look for the maximum number of steps per interval.

```{r}
maxValue <- max(stepsPerInterval$steps)
maxInterval <- stepsPerInterval[stepsPerInterval$steps == maxValue,]$interval
maxSteps <- stepsPerInterval[stepsPerInterval$steps == maxValue,]$steps
```

The interval which has the maximum average number of steps over days is `r maxInterval`, with average number of steps `r maxSteps`.

## Imputing missing values

Calculating the number of NAs in the original dataset:

```{r}
sum(is.na(data$steps))
```

In order to fill all the missing values of the dataset, the proposed approach is, for a given interval, to fill it with the mean for that interval over days.  
The avg function returns the average number of steps over days for a given time interval.

```{r}
avg <- function(interval){
  mean(data[data$interval == interval,]$steps, na.rm = TRUE)
}

newdata <- data
newdata[is.na(newdata$steps),]$steps <- avg(newdata[is.na(newdata$steps),]$interval)

summary(newdata)
```

This time, missing values have been filled in. Calculation of the number of steps per day :

```{r}
stepsPerDay2 <- newdata[,c("steps","date")] %>% group_by(date) %>% summarise_each(funs(sum))
stepsPerDay2
```

The histogram of total number of steps taken per day is now:

```{r histogramStepsPerDayWithoutNAs, fig.width=10}
hist(stepsPerDay2$steps, 
     breaks = 10,
     main = "Histogram of total number of steps taken each day, missing values have been filled in",
     xlab = "Total number of steps taken on a day",
     ylab = "Number of cases")
```


```{r}
meanStepsPerDay2 <- mean(stepsPerDay2$steps)
medianStepsPerDay2 <- median(stepsPerDay2$steps)
```

The mean is now `r meanStepsPerDay2`  
The median is now `r medianStepsPerDay2`

Imputing missing values did not have a major impact on previous estimates.

##Are there differences in activity patterns between weekdays and weekends?

In the following code chunk, the variable "day" is created. It symbolizes if the day of the observation is a weekday or a weeken (two-factor variable).

```{r}
newdata$date <- as.Date(newdata$date,format = "%Y-%M-%d")
newdata$day <- weekdays(newdata$date)

weekdays = c("Lundi","Mardi","Mercredi","Jeudi","Vendredi")
weekend = c("Samedi","Dimanche")

newdata$day[newdata$day %in% weekdays] <- "Weekday"
newdata$day[newdata$day %in% weekend] <- "Weekend"

```

This variable makes it simpler to build a plot showing the average number of steps taken for each time interval, averaged across all weekday days or weekend days.

```{r weekdayweekend,message=FALSE}
stepsPerIntervalDay <- newdata[,c("steps","interval","day")] %>% group_by(interval,day) %>% summarise_each(funs(mean))

library(lattice)
xyplot(steps~interval|day, 
  	  data = stepsPerIntervalDay,
  	  type = "l",
      main="Average steps taken per interval, on weekdays and weekends", 
      ylab="Steps", 
  	  xlab="Time interval")
```

